---
title: "How Does the Performance in Different Industries Affecting the US Market During Recession"
subtitle: "Finding correlation between S&P 500 and 5 different industries during 3 famous recessions"
author: 
  - Zecheng Wu
thanks: "Code and data are available at: https://github.com/nostestwu/market_prediction"
date: "`r format(Sys.time(), '%d %B %Y')`"
abstract: "The S&P 500 index is the best overall measurement for the US market in which it is important for investors to determine whether they should keep more cash or invest more in the market. We have used the S&P 500 PE ratio, a ratio that measures the current share price relative to its earnings per share, to find out the time period for 3 recessions in 2002, 2018 and 2020. Then three tables are created which only contain the closing value for the S&P, AAPL(technology), MAR(real estate), SAFM(agriculture), NEM(gold) and XOM(gas) for each recession. The discovery from the result concludes that agriculture does not seem to get affected by recessions from the market and vice versa. In addition, I have created a multiple linear regression model by using the data from last year till this day to see if we can see any signs that the market might crash. We never know when the market will crash, but what we can do is to search for similar patterns from history and avoid our loss in the future."

output:
  bookdown::pdf_document2
toc: FALSE
bibliography: references.bib
---

```{r setup, include=FALSE}
library(tidyverse)
library(ggplot2)

setwd("/Users/nostest/Desktop/Github/market_prediction")

S_P <- readr::read_csv("inputs/data/HistoricalPrices.csv")
AAPL <- readr::read_csv("inputs/data/tech_AAPL.csv")
MAR <- readr::read_csv("inputs/data/house_MAR.csv")
SAFM <- readr::read_csv("inputs/data/agri_SAFM.csv")
NEM <- readr::read_csv("inputs/data/gold_NEM.csv")
XOM <- readr::read_csv("inputs/data/gas_XOM.csv")

#Cleaning S&P
S_P$Date <- as.Date(S_P$Date, "%m/%d/%y")
```

#Introduction

Recession has always been the fear for people since 2020 due to the COVID pandemic. Governments have printed more money and tried to save businesses when many of them declared bankruptcy that caused the overall market to go down. The action of printing more money creates inflation, but the price for many fundamentals and food didn't seem to inflate in 2021. Many large investing companies have left their position in the trading market and kept more cash in their hands. This would cause the market to have less cash flow and many people believe that the market will crash eventually. However, nothing happened until 2022 February 24, 2022 when Russia invaded Ukraine and many investors started to trade again. Many sanctions against Russia appeared and the current market is at a very dangerous position right now. Nobody knows when the war will end, and some people believe that this war is a buffer for the recession. Recession is not a bad thing, it gives opportunities for many new companies to startup in their country. Uber, Instagram, WhatsApp, etc that became successful after the recession in 2008(foxbusiness). Thus, we want to know which industries might fall and which industries might rise during a recession. 

We begin with the S&P 500 PE ratio (S&P PE ratio) which simply means a higher PE ratio implies the S&P 500 index is more overvalued. Since the S&P 500 is a U.S index, all the companies which I have selected to compare with the index are also from the U.S. I marked the starting and ending point for the peaks. Surprisingly, the years within these peaks are at the exact same time period as the recessions. Knowing that in 2002 it was the tech bubble and in 2008 it was the housing bubble, I have selected Apple(AAPL) as the representative for the technology industry and Marriott International(MAR) for the real estate industry. The sanctions on Russia have largely affected the gas price worldwide recently, thus I have added the Exxon Mobil Corporation(XOM) for the gas industry. Sanderson Farms(SAFM) and Newmont Corporation (NEM) are the representatives for agriculture and gold. I collected these data from yahoo finance starting from 1990 January 1 to 2022 April 21 by daily and insert them into R studio[@citeR]. The first table combines the closing value for S&P 500 and these industry representatives from 2001 January 01 to 2002 september 30. This is because the first peak for the S&P 500 PE ratio is between 2001 first quarter to 2002 third quarter. Next, I have created a multiple linear model setting the S_P as the response variable and five other industries as the explanatory variables. Lastly, I use the summary function to see which variables are significant to our model. I have repeated the same process for the “Great Recession” from 2008/07/01 to 2010/01/01 and the “Covid Recession” from 2020/01/01 to 2021/09/30. In both 2002 and 2008 recession model summary, the agriculture industry does not appear to be significant meaning that this industry is not correlated to the market. This concludes that our best fit model will only contain the values from four industries.

TThe predicting model I have created starts from 2021/04/01 to 2022/04/02 to test out whether our previous assumption was reliable. The exact same process was done except I have created 8 different models to see which one would be the best one to use for our current date. The result shows model 6 was the best model fit, which somehow it is the exact model that we got from the previous three recessions. These findings make me wonder if we know the performance for these four industries, does it mean that we can see the signs for recession? Before we want to predict these stocks, we want to see if the S&P 500 PE ratio has met our criteria. According to the PE ratio chart, every recession has hit above 40 at least. Currently we are at 33.1 right now according to the data from Current Market Valuation on April 22, 2022, which means a recession could potentially happen soon. 

The reason why this model is important is because most people want to have a brief idea about the market and how it is working but don’t know where to start. By understanding the model, we can also think about our society logically. If everyone is losing money in the market, people will still spend money on food even if everything is inflated. The sanctions for the Russian War had a huge impact on gas and energy, but people will still pay more for them since these are the essential things they need to survive just like food. Thus, peace is the best solution for the economy. 

The remainder of this paper is : Section \@ref(data) explains the data that we collected from yahoo finance.Section \@ref(model) shows the model that we are using. Section \@ref(result) explains what we found from the models. Section \@ref(discussion) explains interesting discoveries and future improvements.

# Data

```{r echo=FALSE, warning=FALSE, message=FALSE}


S_P_graph <- ggplot(data = S_P, aes(x = Date, y = Close)) + 
  geom_line() + 
  geom_vline(xintercept = as.numeric(as.Date("2001/01/01")), color = 'red', linetype = 2) +
  geom_vline(xintercept = as.numeric(as.Date("2002/09/30")), color = 'red', linetype = 2) + 
  geom_vline(xintercept = as.numeric(as.Date("2008/07/01")), color = 'red', linetype = 2) + 
  geom_vline(xintercept = as.numeric(as.Date("2010/06/30")), color = 'red', linetype = 2) + 
  geom_vline(xintercept = as.numeric(as.Date("2020/01/01")), color = 'red', linetype = 2) + 
  geom_vline(xintercept = as.numeric(as.Date("2021/09/30")), color = 'red', linetype = 2) +
  labs(title = "S&P 500", caption = "Figure 1")


AAPL_graph <- ggplot(data = AAPL, aes(x = Date, y = Close)) + 
  geom_line() + 
  geom_vline(xintercept = as.numeric(as.Date("2001/01/01")), color = 'red', linetype = 2) +
  geom_vline(xintercept = as.numeric(as.Date("2002/09/30")), color = 'red', linetype = 2) + 
  geom_vline(xintercept = as.numeric(as.Date("2008/07/01")), color = 'red', linetype = 2) + 
  geom_vline(xintercept = as.numeric(as.Date("2010/06/30")), color = 'red', linetype = 2) + 
  geom_vline(xintercept = as.numeric(as.Date("2020/01/01")), color = 'red', linetype = 2) + 
  geom_vline(xintercept = as.numeric(as.Date("2021/09/30")), color = 'red', linetype = 2) +
  labs(title = "APPLE (AAPL)", caption = "Figure 2")


MAR_graph <- ggplot(data = MAR, aes(x = Date, y = Close)) + 
  geom_line() + 
  geom_vline(xintercept = as.numeric(as.Date("2001/01/01")), color = 'red', linetype = 2) +
  geom_vline(xintercept = as.numeric(as.Date("2002/09/30")), color = 'red', linetype = 2) + 
  geom_vline(xintercept = as.numeric(as.Date("2008/07/01")), color = 'red', linetype = 2) + 
  geom_vline(xintercept = as.numeric(as.Date("2010/06/30")), color = 'red', linetype = 2) + 
  geom_vline(xintercept = as.numeric(as.Date("2020/01/01")), color = 'red', linetype = 2) + 
  geom_vline(xintercept = as.numeric(as.Date("2021/09/30")), color = 'red', linetype = 2) +
  labs(title = "Marriott International (MAR)", caption = "Figure 3")

SAFM_graph <- ggplot(data = SAFM, aes(x = Date, y = Close)) + 
  geom_line() + 
  geom_vline(xintercept = as.numeric(as.Date("2001/01/01")), color = 'red', linetype = 2) +
  geom_vline(xintercept = as.numeric(as.Date("2002/09/30")), color = 'red', linetype = 2) + 
  geom_vline(xintercept = as.numeric(as.Date("2008/07/01")), color = 'red', linetype = 2) + 
  geom_vline(xintercept = as.numeric(as.Date("2010/06/30")), color = 'red', linetype = 2) + 
  geom_vline(xintercept = as.numeric(as.Date("2020/01/01")), color = 'red', linetype = 2) + 
  geom_vline(xintercept = as.numeric(as.Date("2021/09/30")), color = 'red', linetype = 2) +
  labs(title = "Sanderson Farms, Inc. (SAFM)", caption = "Figure 4")

NEM_graph <- ggplot(data = NEM, aes(x = Date, y = Close)) + 
  geom_line() + 
  geom_vline(xintercept = as.numeric(as.Date("2001/01/01")), color = 'red', linetype = 2) +
  geom_vline(xintercept = as.numeric(as.Date("2002/09/30")), color = 'red', linetype = 2) + 
  geom_vline(xintercept = as.numeric(as.Date("2008/07/01")), color = 'red', linetype = 2) + 
  geom_vline(xintercept = as.numeric(as.Date("2010/06/30")), color = 'red', linetype = 2) + 
  geom_vline(xintercept = as.numeric(as.Date("2020/01/01")), color = 'red', linetype = 2) + 
  geom_vline(xintercept = as.numeric(as.Date("2021/09/30")), color = 'red', linetype = 2) +
  labs(title = "Newmont Corporation (NEM)", caption = "Figure 5")

XOM_graph <- ggplot(data = XOM, aes(x = Date, y = Close)) + 
  geom_line() + 
  geom_vline(xintercept = as.numeric(as.Date("2001/01/01")), color = 'red', linetype = 2) +
  geom_vline(xintercept = as.numeric(as.Date("2002/09/30")), color = 'red', linetype = 2) + 
  geom_vline(xintercept = as.numeric(as.Date("2008/07/01")), color = 'red', linetype = 2) + 
  geom_vline(xintercept = as.numeric(as.Date("2010/06/30")), color = 'red', linetype = 2) + 
  geom_vline(xintercept = as.numeric(as.Date("2020/01/01")), color = 'red', linetype = 2) + 
  geom_vline(xintercept = as.numeric(as.Date("2021/09/30")), color = 'red', linetype = 2) +
  labs(title = "Exxon Mobil Corporation (XOM)", caption = "Figure 6")

S_P_graph
AAPL_graph
MAR_graph
SAFM_graph
NEM_graph
XOM_graph

```


S&P 500 index tracks 500 publicly traded domestic companies in the U.S and the most well known index for people to use and see the overall U.S market. PE ratio stands for price per earning ratio, which means how many times the company is trading its earnings per share. For instance, a company has an earning of $100000 and 50000 shares outstanding. This means the earnings per share is $2 meaning each share you hold worth $2 dollars of its earnings. If this company offers each share $100, this means the PE ratio is 50, meaning this company is giving out the shares 50 times higher than its earning. It is obvious to see that we want to invest in companies that sell their shares very cheap and the earnings are super high which have a low PE ratio. Therefore, this is the reason why I consider the high PE ratio in S&P 500 to see when companies have their shares at a super high price while they are not making much money. 

The five companies I have chosen to compare with the index are Apple(AAPL), Marriott International(MAR), Sanderson Farms(SAFM), Newmont Corporation (NEM), and Exxon Mobil Corporation(XOM). The reason why I have selected these companies is because all of these companies are one of the largest in their own industry and they have data since 1990. The first step I did is to download the daily data for each of these companies from yahoo finance and import them into rstudio. Then I went to the S&P 500 PE ratio website to record the time interval for each PE ratio peak. Unfortunately I do not have the account else I would have downloaded the actual data from the website. Then I have created a line graph for each of the companies and the index to see a general trend from 1990 to 2022 as we can see from Figure 1 to Figure 6. Instead of considering the entire chart, we want to break them into smaller parts which is the PE ratio time intervals we found earlier. I have combined the closed values from each company, renaming them and creating a model having the S&P as the y variable and different companies as the x variables. The first recession period starts from 2001/01/01 to 2002/09/30. I have repeated the same procedure for intervals from 2008/07/01 to 2010/06/30 and from 2020/01/01 to 2021/09/30, the summary tables are displayed as following:


```{r echo=FALSE, warning=FALSE, message=FALSE}

# Recession 1: tech bubble

r1_S_P <- S_P %>% filter(Date >= as.Date("2001/01/01") & Date <= as.Date("2002/09/30"))
r1_AAPL <- AAPL %>% filter(Date >= as.Date("2001/01/01") & Date <= as.Date("2002/09/30"))
r1_MAR <- MAR %>% filter(Date >= as.Date("2001/01/01") & Date <= as.Date("2002/09/30"))
r1_SAFM <- SAFM %>% filter(Date >= as.Date("2001/01/01") & Date <= as.Date("2002/09/30"))
r1_NEM <- NEM %>% filter(Date >= as.Date("2001/01/01") & Date <= as.Date("2002/09/30"))
r1_XOM <- XOM %>% filter(Date >= as.Date("2001/01/01") & Date <= as.Date("2002/09/30"))

#combining all tables

r1_S_P <- rename(r1_S_P, S_P = Close) %>% 
  select(Date, S_P)
r1_AAPL <- rename(r1_AAPL, AAPL = Close) %>% 
  select(Date, AAPL)
r1_MAR <- rename(r1_MAR, MAR = Close) %>% 
  select(Date, MAR)
r1_SAFM <- rename(r1_SAFM, SAFM = Close) %>% 
  select(Date, SAFM)
r1_NEM <- rename(r1_NEM, NEM = Close) %>% 
  select(Date, NEM)
r1_XOM <- rename(r1_XOM, XOM = Close) %>% 
  select(Date, XOM)

clean1 <- inner_join(r1_S_P, r1_AAPL, by = "Date")
clean2 <- inner_join(clean1, r1_MAR, by = "Date")
clean3 <- inner_join(clean2, r1_SAFM, by = "Date")
clean4 <- inner_join(clean3, r1_NEM, by = "Date")
recession1 <- inner_join(clean4, r1_XOM, by = "Date")

#creating model

model1 <- lm(S_P ~ AAPL + MAR + SAFM + NEM + XOM , data = recession1)
summary(model1)

# Recession 2: housing bubble

r2_S_P <- S_P %>% filter(Date >= as.Date("2008/07/01") & Date <= as.Date("2010/06/30"))
r2_AAPL <- AAPL %>% filter(Date >= as.Date("2008/07/01") & Date <= as.Date("2010/06/30"))
r2_MAR <- MAR %>% filter(Date >= as.Date("2008/07/01") & Date <= as.Date("2010/06/30"))
r2_SAFM <- SAFM %>% filter(Date >= as.Date("2008/07/01") & Date <= as.Date("2010/06/30"))
r2_NEM <- NEM %>% filter(Date >= as.Date("2008/07/01") & Date <= as.Date("2010/06/30"))
r2_XOM <- XOM %>% filter(Date >= as.Date("2008/07/01") & Date <= as.Date("2010/06/30"))

#combining all tables

r2_S_P <- rename(r2_S_P, S_P = Close) %>% 
  select(Date, S_P)
r2_AAPL <- rename(r2_AAPL, AAPL = Close) %>% 
  select(Date, AAPL)
r2_MAR <- rename(r2_MAR, MAR = Close) %>% 
  select(Date, MAR)
r2_SAFM <- rename(r2_SAFM, SAFM = Close) %>% 
  select(Date, SAFM)
r2_NEM <- rename(r2_NEM, NEM = Close) %>% 
  select(Date, NEM)
r2_XOM <- rename(r2_XOM, XOM = Close) %>% 
  select(Date, XOM)

clean6 <- inner_join(r2_S_P, r2_AAPL, by = "Date")
clean7 <- inner_join(clean6, r2_MAR, by = "Date")
clean8 <- inner_join(clean7, r2_SAFM, by = "Date")
clean9 <- inner_join(clean8, r2_NEM, by = "Date")
recession2 <- inner_join(clean9, r2_XOM, by = "Date")

#creating model

model2 <- lm(S_P ~ AAPL + MAR + SAFM + NEM + XOM , data = recession2)
summary(model2)

# Recession 3: COVID

r3_S_P <- S_P %>% filter(Date >= as.Date("2020/01/01") & Date <= as.Date("2021/09/30"))
r3_AAPL <- AAPL %>% filter(Date >= as.Date("2020/01/01") & Date <= as.Date("2021/09/30"))
r3_MAR <- MAR %>% filter(Date >= as.Date("2020/01/01") & Date <= as.Date("2021/09/30"))
r3_SAFM <- SAFM %>% filter(Date >= as.Date("2020/01/01") & Date <= as.Date("2021/09/30"))
r3_NEM <- NEM %>% filter(Date >= as.Date("2020/01/01") & Date <= as.Date("2021/09/30"))
r3_XOM <- XOM %>% filter(Date >= as.Date("2020/01/01") & Date <= as.Date("2021/09/30"))

#combining all tables

r3_S_P <- rename(r3_S_P, S_P = Close) %>% 
  select(Date, S_P)
r3_AAPL <- rename(r3_AAPL, AAPL = Close) %>% 
  select(Date, AAPL)
r3_MAR <- rename(r3_MAR, MAR = Close) %>% 
  select(Date, MAR)
r3_SAFM <- rename(r3_SAFM, SAFM = Close) %>% 
  select(Date, SAFM)
r3_NEM <- rename(r3_NEM, NEM = Close) %>% 
  select(Date, NEM)
r3_XOM <- rename(r3_XOM, XOM = Close) %>% 
  select(Date, XOM)

clean11 <- inner_join(r3_S_P, r3_AAPL, by = "Date")
clean12 <- inner_join(clean11, r3_MAR, by = "Date")
clean13 <- inner_join(clean12, r3_SAFM, by = "Date")
clean14 <- inner_join(clean13, r3_NEM, by = "Date")
recession3 <- inner_join(clean14, r3_XOM, by = "Date")

#creating model

model3 <- lm(S_P ~ AAPL + MAR + SAFM + NEM + XOM , data = recession3)
summary(model3)

```

If we only consider the summary for the first and second recession, we can clearly see that SAFM is always the least significant variable for our model. Even though in recession 1, the P value for SAFM is very significant. However, the P value it has is relatively greater than all other variables. Thus, we don’t want to include this variable in our prediction model.

After finding out the variables that are significant to determine the market, wouldn’t it be more interesting if we can predict the future market trend? If we look closely at the time period for these recessions, we can see that they are around 1 year and 9 months to 2 years. Thus, I decided to take the data within a year from 2021/04/01 to 2022/04/01 and predict the market in the next 9 months. Simply repeating the same procedures except for the final step which we want to compare different models by keeping different x variables. I have created 8 different models and compared the summary between them. Then I picked 5 of the best summary models to predict the S&P value and compare it with the actual value on 2022/04/07,  2022/04/14 and 2022/04/22 which is a one week interval. I sum up the difference for each week and compare between these 5 models to see which one has the smallest difference. It turns out model6 has the smallest difference which we will be using this model for our prediction.

```{r echo=FALSE, warning=FALSE, message=FALSE}


p_S_P <- S_P %>% filter(Date >= as.Date("2021/04/01") & Date <= as.Date("2022/04/01"))
p_AAPL <- AAPL %>% filter(Date >= as.Date("2021/04/01") & Date <= as.Date("2022/04/01"))
p_MAR <- MAR %>% filter(Date >= as.Date("2021/04/01") & Date <= as.Date("2022/04/01"))
p_SAFM <- SAFM %>% filter(Date >= as.Date("2021/04/01") & Date <= as.Date("2022/04/01"))
p_NEM <- NEM %>% filter(Date >= as.Date("2021/04/01") & Date <= as.Date("2022/04/01"))
p_XOM <- XOM %>% filter(Date >= as.Date("2021/04/01") & Date <= as.Date("2022/04/01"))

#combining all tables

p_S_P <- rename(p_S_P, S_P = Close) %>% 
  select(Date, S_P)
p_AAPL <- rename(p_AAPL, AAPL = Close) %>% 
  select(Date, AAPL)
p_MAR <- rename(p_MAR, MAR = Close) %>% 
  select(Date, MAR)
p_SAFM <- rename(p_SAFM, SAFM = Close) %>% 
  select(Date, SAFM)
p_NEM <- rename(p_NEM, NEM = Close) %>% 
  select(Date, NEM)
p_XOM <- rename(p_XOM, XOM = Close) %>% 
  select(Date, XOM)

clean16 <- inner_join(p_S_P, p_AAPL, by = "Date")
clean17 <- inner_join(clean16, p_MAR, by = "Date")
clean18 <- inner_join(clean17, p_SAFM, by = "Date")
clean19 <- inner_join(clean18, p_NEM, by = "Date")
clean20 <- inner_join(clean19, p_XOM, by = "Date")

#creating model

model4 <- lm(S_P ~ AAPL + MAR + SAFM + NEM + XOM , data = clean20)
model5 <- lm(S_P ~ MAR + NEM + XOM , data = clean20)
model6 <- lm(S_P ~ AAPL + MAR + NEM + XOM , data = clean20)
model7 <- lm(S_P ~ AAPL + MAR + NEM , data = clean20)
model8 <- lm(S_P ~ AAPL + MAR + SAFM, data = clean20)
model9 <- lm(S_P ~ AAPL + MAR + SAFM + NEM , data = clean20)
model10 <- lm(S_P ~ AAPL + MAR + SAFM + XOM , data = clean20)
model11 <- lm(S_P ~ AAPL + NEM , data = clean20)

library(knitr)

t <- data.frame(date = c('2022/04/21','2022/04/14', '2022/04/7'), model5 = c(4453.65315, 4301.77435,  4256.54238), model6 = c(4398.247926, 4327.545053,  4399.516128), model7 = c(4399.393657, 4310.553853,  4409.573669), model10 = c(4521.964031, 4491.661995,  4468.331013), model11 = c(4401.447712, 4311.831264,  4408.804913), Actual_S_P = c(4393.66, 4392.59,  4500.21))

kable(t, caption = "Model Prediction")

```

# Model

The model that we will be using is the multiple linear regression model. We will write the model as the following: 

$$ \bar Y = \beta_0 + \beta_1X_1 + \beta_1X_2 + \beta_1X_3 + \beta_1X_4 $$
$\bar Y$: the estimated S&P 500 index value

$\beta_0$: intercept

$\beta_1 ... \beta_n$: coefficents for $X_1$ to $X_n$ 

$X_1$: AAPL value from model6 (9.686)

$X_2$: MAR value from model6 (2.2753)

$X_3$: NEM value from model6 (-5.9445)

$X_4$: XOM value from model6 (-6.9975)

# Result

If we compute the sum of the differences for all the models from Table 1, we would get -274.49 for model 5, -161.15 for model 6, -166.94 for model 7, 195.50 for model 10 and -164.38 for model 11. We can clearly see that model 6, model 7 and model 11 have similar values. Model 6 has 4 variables, model 7 has 3 variables and model 11 has 2 variables. Normally we would choose the model that has the least variable which is model 11. Considering the previous 3 recessions where each one of these variables are significant, the summary table in mode 6 shows that all variables are significant for our model.

Model 6 has positive coefficients for AAPL and MAR and negative coefficients for NEM and XOM. This is not something ideal to see because the gas price is increasing significantly after the Russian war started. If we see the exports from Russia, gold is also one of the top exports according to the OEC website (OEC) Sanctions on Russia could potentially raise the stock prices for both gas and gold industries since there would be less resources in the country without the imports from Russia. Our model shows that there is a negative correlation for both gas and gold with the S&P 500 index, a rapid increase of gas and gold price could lead to a rapid decrease in the overall market. In addition, many technology companies have moved out from Russia. This reduces the number of Russian customers, causing the price for technology companies to go down. Knowing that technology has a positive correlation with the index, decrease in technology will lead to a decrease in the overall market. This seems like the action of sanction for the U.S is speeding up the time for a new recession. 

All in all, many signs are indicating that the overall market is going down. The PE ratio has risen from the 2021 fourth quarter to 2022 first quarter indicating a rise in the overvalued S&P Index. The sanctions on Russia have further damaged the overall market. Luckly, many countries still buy resources from Russia undercover that slows the rise in gas and gold price. However, we are only considering the U.S market which most places are already suffering from the rise in gas price. Therefore, we conclude that a recession might happen within one to four quarters after the PE ratio breaks the 40 mark.

```{r echo=FALSE, warning=FALSE, message=FALSE}


summary(model6)

```

# Discussion

## First discussion point

If we observe the graph for each company and the index, we can see that in 2015 the overall market dropped without any sign. After researching, it appears to be a market selloff during 2015 to 2016. This means even if the PE ratio for S&P doesn’t rise, the overall market could still crash if all the investors are panicking about the current market. This could be the reason that explains why governments are printing more money to save the current market because everyone was too scared to invest after the COVID recession in 2020. This human made market crash is more unpredictable compared to a bubble market crash where there is no sign on the PE ratio chart. Many countries have already printed a lot of money to save the current market. What would happen if a bubble market and a human made market both happened at the same time? This could be the time that we all need to be aware of where we invest our money in.


## Second discussion point

After researching the 505 companies within the S&P 500, it turns out all the companies except for SAFM are included. This is why all companies have a strong correlation with the index except for SAFM. Even though this is a bad company selection, the purpose of this paper is to find the patterns during recession periods. Our goal is to see if different industries have any signs or correlations with the overall market trend and see if we can use these patterns to predict the next recession. In addition, we also want to know whether the variable is positively or negatively correlated to the S&P index. One of the methods that can improve this is to choose more companies from the same industry but not within the index. It would allow us to diversify the selections within each industry.

## Third discussion point

In 2002 it was the tech bubble and 2008 was the real estate bubble. Both of these recessions appear to have technology and real estate being positively correlated where gold and gas are negatively correlated. This means buying gold and gas during that time would make a lot of money. However, the COVID 2020 recession has everything positively correlated which means you will lose money no matter what you invest in the market. Currently, our result shows that there is an indication for a recession which is similar in 2002 and 2008. The issue is that the pandemic isn’t over yet which means we should think about whether it is still right to invest money in gas and gold. Even though our model shows they are negatively correlated with the index, we need to understand that pandemic hits the market entirely and this market fall is not caused by the bubble.


## Weaknesses and next steps

One major weakness is that we only selected one representative from each industry to conclude everything. If we can use more than 5 companies from each industry, then the data is more reliable to represent the entire industry growth. If we can compare the data with the index from other countries, then there might be some interesting observations that we can make. I had a quick glance at the SSE Index in China and the FTSE Index in London, both of them seem to have a rapid squeeze to the index and a rapid fall. Unlike the S&P where it falls rapidly at start but recovers back at a similar pace.Thus, comparing the data worldwide could find out more correlations and signs before each recession. 


\newpage

\appendix

# Appendix {-}

```{r}

setwd("/Users/nostest/Desktop/Github/market_prediction")
library(tidyverse)
library(ggplot2)

# Read in the raw data. 

S_P <- readr::read_csv("inputs/data/HistoricalPrices.csv")
AAPL <- readr::read_csv("inputs/data/tech_AAPL.csv")
MAR <- readr::read_csv("inputs/data/house_MAR.csv")
SAFM <- readr::read_csv("inputs/data/agri_SAFM.csv")
NEM <- readr::read_csv("inputs/data/gold_NEM.csv")
XOM <- readr::read_csv("inputs/data/gas_XOM.csv")

#Cleaning S&P
S_P$Date <- as.Date(S_P$Date, "%m/%d/%y")

#### What's next? ####

S_P_graph <- ggplot(data = S_P, aes(x = Date, y = Close)) + 
  geom_line() + 
  geom_vline(xintercept = as.numeric(as.Date("2001/01/01")), color = 'red', linetype = 2) +
  geom_vline(xintercept = as.numeric(as.Date("2002/09/30")), color = 'red', linetype = 2) + 
  geom_vline(xintercept = as.numeric(as.Date("2008/07/01")), color = 'red', linetype = 2) + 
  geom_vline(xintercept = as.numeric(as.Date("2010/06/30")), color = 'red', linetype = 2) + 
  geom_vline(xintercept = as.numeric(as.Date("2020/01/01")), color = 'red', linetype = 2) + 
  geom_vline(xintercept = as.numeric(as.Date("2021/09/30")), color = 'red', linetype = 2) 


AAPL_graph <- ggplot(data = AAPL, aes(x = Date, y = Close)) + 
  geom_line() + 
  geom_vline(xintercept = as.numeric(as.Date("2001/01/01")), color = 'red', linetype = 2) +
  geom_vline(xintercept = as.numeric(as.Date("2002/09/30")), color = 'red', linetype = 2) + 
  geom_vline(xintercept = as.numeric(as.Date("2008/07/01")), color = 'red', linetype = 2) + 
  geom_vline(xintercept = as.numeric(as.Date("2010/06/30")), color = 'red', linetype = 2) + 
  geom_vline(xintercept = as.numeric(as.Date("2020/01/01")), color = 'red', linetype = 2) + 
  geom_vline(xintercept = as.numeric(as.Date("2021/09/30")), color = 'red', linetype = 2) 

MAR_graph <- ggplot(data = MAR, aes(x = Date, y = Close)) + 
  geom_line() + 
  geom_vline(xintercept = as.numeric(as.Date("2001/01/01")), color = 'red', linetype = 2) +
  geom_vline(xintercept = as.numeric(as.Date("2002/09/30")), color = 'red', linetype = 2) + 
  geom_vline(xintercept = as.numeric(as.Date("2008/07/01")), color = 'red', linetype = 2) + 
  geom_vline(xintercept = as.numeric(as.Date("2010/06/30")), color = 'red', linetype = 2) + 
  geom_vline(xintercept = as.numeric(as.Date("2020/01/01")), color = 'red', linetype = 2) + 
  geom_vline(xintercept = as.numeric(as.Date("2021/09/30")), color = 'red', linetype = 2) 

SAFM_graph <- ggplot(data = SAFM, aes(x = Date, y = Close)) + 
  geom_line() + 
  geom_vline(xintercept = as.numeric(as.Date("2001/01/01")), color = 'red', linetype = 2) +
  geom_vline(xintercept = as.numeric(as.Date("2002/09/30")), color = 'red', linetype = 2) + 
  geom_vline(xintercept = as.numeric(as.Date("2008/07/01")), color = 'red', linetype = 2) + 
  geom_vline(xintercept = as.numeric(as.Date("2010/06/30")), color = 'red', linetype = 2) + 
  geom_vline(xintercept = as.numeric(as.Date("2020/01/01")), color = 'red', linetype = 2) + 
  geom_vline(xintercept = as.numeric(as.Date("2021/09/30")), color = 'red', linetype = 2) 

NEM_graph <- ggplot(data = NEM, aes(x = Date, y = Close)) + 
  geom_line() + 
  geom_vline(xintercept = as.numeric(as.Date("2001/01/01")), color = 'red', linetype = 2) +
  geom_vline(xintercept = as.numeric(as.Date("2002/09/30")), color = 'red', linetype = 2) + 
  geom_vline(xintercept = as.numeric(as.Date("2008/07/01")), color = 'red', linetype = 2) + 
  geom_vline(xintercept = as.numeric(as.Date("2010/06/30")), color = 'red', linetype = 2) + 
  geom_vline(xintercept = as.numeric(as.Date("2020/01/01")), color = 'red', linetype = 2) + 
  geom_vline(xintercept = as.numeric(as.Date("2021/09/30")), color = 'red', linetype = 2) 

XOM_graph <- ggplot(data = XOM, aes(x = Date, y = Close)) + 
  geom_line() + 
  geom_vline(xintercept = as.numeric(as.Date("2001/01/01")), color = 'red', linetype = 2) +
  geom_vline(xintercept = as.numeric(as.Date("2002/09/30")), color = 'red', linetype = 2) + 
  geom_vline(xintercept = as.numeric(as.Date("2008/07/01")), color = 'red', linetype = 2) + 
  geom_vline(xintercept = as.numeric(as.Date("2010/06/30")), color = 'red', linetype = 2) + 
  geom_vline(xintercept = as.numeric(as.Date("2020/01/01")), color = 'red', linetype = 2) + 
  geom_vline(xintercept = as.numeric(as.Date("2021/09/30")), color = 'red', linetype = 2) 


# Recession 1: tech bubble

r1_S_P <- S_P %>% filter(Date >= as.Date("2001/01/01") & Date <= as.Date("2002/09/30"))
r1_AAPL <- AAPL %>% filter(Date >= as.Date("2001/01/01") & Date <= as.Date("2002/09/30"))
r1_MAR <- MAR %>% filter(Date >= as.Date("2001/01/01") & Date <= as.Date("2002/09/30"))
r1_SAFM <- SAFM %>% filter(Date >= as.Date("2001/01/01") & Date <= as.Date("2002/09/30"))
r1_NEM <- NEM %>% filter(Date >= as.Date("2001/01/01") & Date <= as.Date("2002/09/30"))
r1_XOM <- XOM %>% filter(Date >= as.Date("2001/01/01") & Date <= as.Date("2002/09/30"))

#combining all tables

r1_S_P <- rename(r1_S_P, S_P = Close) %>% 
  select(Date, S_P)
r1_AAPL <- rename(r1_AAPL, AAPL = Close) %>% 
  select(Date, AAPL)
r1_MAR <- rename(r1_MAR, MAR = Close) %>% 
  select(Date, MAR)
r1_SAFM <- rename(r1_SAFM, SAFM = Close) %>% 
  select(Date, SAFM)
r1_NEM <- rename(r1_NEM, NEM = Close) %>% 
  select(Date, NEM)
r1_XOM <- rename(r1_XOM, XOM = Close) %>% 
  select(Date, XOM)

clean1 <- inner_join(r1_S_P, r1_AAPL, by = "Date")
clean2 <- inner_join(clean1, r1_MAR, by = "Date")
clean3 <- inner_join(clean2, r1_SAFM, by = "Date")
clean4 <- inner_join(clean3, r1_NEM, by = "Date")
clean5 <- inner_join(clean4, r1_XOM, by = "Date")

#creating model

model1 <- lm(S_P ~ AAPL + MAR + SAFM + NEM + XOM , data = clean5)

# Recession 2: housing bubble

r2_S_P <- S_P %>% filter(Date >= as.Date("2008/07/01") & Date <= as.Date("2010/06/30"))
r2_AAPL <- AAPL %>% filter(Date >= as.Date("2008/07/01") & Date <= as.Date("2010/06/30"))
r2_MAR <- MAR %>% filter(Date >= as.Date("2008/07/01") & Date <= as.Date("2010/06/30"))
r2_SAFM <- SAFM %>% filter(Date >= as.Date("2008/07/01") & Date <= as.Date("2010/06/30"))
r2_NEM <- NEM %>% filter(Date >= as.Date("2008/07/01") & Date <= as.Date("2010/06/30"))
r2_XOM <- XOM %>% filter(Date >= as.Date("2008/07/01") & Date <= as.Date("2010/06/30"))

#combining all tables

r2_S_P <- rename(r2_S_P, S_P = Close) %>% 
  select(Date, S_P)
r2_AAPL <- rename(r2_AAPL, AAPL = Close) %>% 
  select(Date, AAPL)
r2_MAR <- rename(r2_MAR, MAR = Close) %>% 
  select(Date, MAR)
r2_SAFM <- rename(r2_SAFM, SAFM = Close) %>% 
  select(Date, SAFM)
r2_NEM <- rename(r2_NEM, NEM = Close) %>% 
  select(Date, NEM)
r2_XOM <- rename(r2_XOM, XOM = Close) %>% 
  select(Date, XOM)

clean6 <- inner_join(r2_S_P, r2_AAPL, by = "Date")
clean7 <- inner_join(clean6, r2_MAR, by = "Date")
clean8 <- inner_join(clean7, r2_SAFM, by = "Date")
clean9 <- inner_join(clean8, r2_NEM, by = "Date")
clean10 <- inner_join(clean9, r2_XOM, by = "Date")

#creating model

model2 <- lm(S_P ~ AAPL + MAR + SAFM + NEM + XOM , data = clean10)

# Recession 3: COVID

r3_S_P <- S_P %>% filter(Date >= as.Date("2020/01/01") & Date <= as.Date("2021/09/30"))
r3_AAPL <- AAPL %>% filter(Date >= as.Date("2020/01/01") & Date <= as.Date("2021/09/30"))
r3_MAR <- MAR %>% filter(Date >= as.Date("2020/01/01") & Date <= as.Date("2021/09/30"))
r3_SAFM <- SAFM %>% filter(Date >= as.Date("2020/01/01") & Date <= as.Date("2021/09/30"))
r3_NEM <- NEM %>% filter(Date >= as.Date("2020/01/01") & Date <= as.Date("2021/09/30"))
r3_XOM <- XOM %>% filter(Date >= as.Date("2020/01/01") & Date <= as.Date("2021/09/30"))

#combining all tables

r3_S_P <- rename(r3_S_P, S_P = Close) %>% 
  select(Date, S_P)
r3_AAPL <- rename(r3_AAPL, AAPL = Close) %>% 
  select(Date, AAPL)
r3_MAR <- rename(r3_MAR, MAR = Close) %>% 
  select(Date, MAR)
r3_SAFM <- rename(r3_SAFM, SAFM = Close) %>% 
  select(Date, SAFM)
r3_NEM <- rename(r3_NEM, NEM = Close) %>% 
  select(Date, NEM)
r3_XOM <- rename(r3_XOM, XOM = Close) %>% 
  select(Date, XOM)

clean11 <- inner_join(r3_S_P, r3_AAPL, by = "Date")
clean12 <- inner_join(clean11, r3_MAR, by = "Date")
clean13 <- inner_join(clean12, r3_SAFM, by = "Date")
clean14 <- inner_join(clean13, r3_NEM, by = "Date")
clean15 <- inner_join(clean14, r3_XOM, by = "Date")

#creating model

model3 <- lm(S_P ~ AAPL + MAR + SAFM + NEM + XOM , data = clean15)


# Predicting model

p_S_P <- S_P %>% filter(Date >= as.Date("2021/04/01") & Date <= as.Date("2022/04/01"))
p_AAPL <- AAPL %>% filter(Date >= as.Date("2021/04/01") & Date <= as.Date("2022/04/01"))
p_MAR <- MAR %>% filter(Date >= as.Date("2021/04/01") & Date <= as.Date("2022/04/01"))
p_SAFM <- SAFM %>% filter(Date >= as.Date("2021/04/01") & Date <= as.Date("2022/04/01"))
p_NEM <- NEM %>% filter(Date >= as.Date("2021/04/01") & Date <= as.Date("2022/04/01"))
p_XOM <- XOM %>% filter(Date >= as.Date("2021/04/01") & Date <= as.Date("2022/04/01"))

#combining all tables

p_S_P <- rename(p_S_P, S_P = Close) %>% 
  select(Date, S_P)
p_AAPL <- rename(p_AAPL, AAPL = Close) %>% 
  select(Date, AAPL)
p_MAR <- rename(p_MAR, MAR = Close) %>% 
  select(Date, MAR)
p_SAFM <- rename(p_SAFM, SAFM = Close) %>% 
  select(Date, SAFM)
p_NEM <- rename(p_NEM, NEM = Close) %>% 
  select(Date, NEM)
p_XOM <- rename(p_XOM, XOM = Close) %>% 
  select(Date, XOM)

clean16 <- inner_join(p_S_P, p_AAPL, by = "Date")
clean17 <- inner_join(clean16, p_MAR, by = "Date")
clean18 <- inner_join(clean17, p_SAFM, by = "Date")
clean19 <- inner_join(clean18, p_NEM, by = "Date")
clean20 <- inner_join(clean19, p_XOM, by = "Date")

#creating model

model4 <- lm(S_P ~ AAPL + MAR + SAFM + NEM + XOM , data = clean20)
model5 <- lm(S_P ~ MAR + NEM + XOM , data = clean20)
model6 <- lm(S_P ~ AAPL + MAR + NEM + XOM , data = clean20)
model7 <- lm(S_P ~ AAPL + MAR + NEM , data = clean20)
model8 <- lm(S_P ~ AAPL + MAR + SAFM, data = clean20)
model9 <- lm(S_P ~ AAPL + MAR + SAFM + NEM , data = clean20)
model10 <- lm(S_P ~ AAPL + MAR + SAFM + XOM , data = clean20)
model11 <- lm(S_P ~ AAPL + NEM , data = clean20)





```


\newpage


# References

Current Market Valuation, https://www.currentmarketvaluation.com/models/price-earnings.php#:~:text=The%20current%20S%26P500%2010%2Dyear%20P%2FE%20Ratio%20is%2033.1.

foxbusiness, https://www.foxbusiness.com/markets/startups-great-recession

Recessions history, https://en.wikipedia.org/wiki/List_of_recessions_in_the_United_States

Russia resources, 
https://oec.world/en/profile/country/rus#:~:text=Exports%20The%20top%20exports%20of,and%20Germany%20(%2414.2B)

Russian sanctions, https://graphics.reuters.com/UKRAINE-CRISIS/SANCTIONS/byvrjenzmve/

S&P 500 Historical Data, https://www.wsj.com/market-data/quotes/index/SPX/historical-prices

S&P 500 PE ratio, https://ycharts.com/indicators/sp_500_pe_ratio

Apple (AAPL), https://finance.yahoo.com/quote/AAPL/history?p=AAPL

marriott international(MAR), https://finance.yahoo.com/quote/MAR/

Sanderson Farms, Inc. (SAFM), https://finance.yahoo.com/quote/SAFM/history?p=SAFM

Newmont Corporation (NEM), https://finance.yahoo.com/quote/NEM/

Exxon Mobil Corporation (XOM), https://finance.yahoo.com/quote/XOM/



